Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-20 demystify.tex
updates
Updates from Overleaf

Here is some general advice for finding a prof.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that prof. and contact them (\autoref{sec:contact}).

@@

Another effective way is taking graduate-level courses in the topics you are interested in.  Profs teaching \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof. after classes, etc---being stand out.  Many profs, including myself, prefer taking in new students this way.  It gives both the prof and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).

You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I do this with multiple students, many of whom are undergrads. Many will drop out because they find they don't like my research, but some will stay and become my PhD students.

@@

\paragraph{Red flags} Various types of red flags can cause concern to the adcom (e.g., plagiarism, cheating, or other academic dishonesty). Another one is that you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.

@@

PhD students often ask why their salary is so low compared to the large grants their advisers get. They also wonder why their offer letters sometimes show that their benefits are higher than what they receive (i.e., stipend).

@@

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

@@

\paragraph{School \autoref{sec:your-school}} We already know about your school, which is a plus (see \autoref{sec:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive and can significantly improve your admission chances if you get them.

@@

I've seen many students, especially international, face a dilemma when they \emph{commit} to a graduate offer but then receive a better one. Advice given in online forums is often along the line that it's okay to switch, using reasons like you haven't yet had a strong relationship with the prof. or you should prioritize personal benefit.

@@

However, regardless of the resolution, retracting an acceptance can have ethical implications. When you accept an offer, you are committing to work with that prof, who then might stop looking for other students. So by retracting your offer, you are breaking your commitment and also causing a great deal of inconvenience to the prof and also taking away the opportunity from other students. 
Ultimately, this choice is personal, involving a balance between personal benefit and the ethical impact on others.

@@

\section{Costly Gifts}\label{sec:gifts} In many countries, it is customary to give professors costly gifts (e.g., expensive liquors, jewelry, or even money), often during the holidays, to show respect and appreciation.  In the US, this is can be considered \emph{inappropriate} and discouraged.

>>>>

Here is some advice for finding a prof.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that prof. and contact them (\autoref{sec:contact}).

@@

Another way is taking graduate-level courses in the topics you are interested in.  Many profs teach \emph{special topics courses} and \emph{research seminars}, and they might be recruiting students. Do well in the class, answer questions, talk to the prof. after classes, etc---being stand out.  Many profs, including myself, prefer taking in new students this way.  It gives both the prof and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds like a marriage!).

You can also ask to do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I do this with several students, some of whom are undergrads. Many will drop out because they find they don't like my research, but some find that they like the work.

@@

\paragraph{Red flags} Various types of red flags can cause concern to the adcom (e.g., plagiarism, cheating, or other academic dishonesty). Another one is that you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). Adcom members might have contacts in other programs and heard about your stories, or your LoRs might mention these.

@@

PhD students often ask why their salary is low compared to the large grants their advisers get. They also wonder why their offer letters sometimes show that their benefits are higher than what they receive (i.e., stipend).

@@

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.

@@

\paragraph{School \autoref{sec:your-school}} Adcom already knows about your school, which is a plus (see \autoref{sec:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive and can significantly improve your admission chances.

@@

I've seen many students, especially international, face a dilemma when they \emph{commit} to a graduate offer but then receive a better one. Advice given in online forums is often along the line that it's okay to switch, using reasons like you haven't yet had a strong relationship with the prof. or you should prioritize your personal benefit.

@@

However, in general, retracting an acceptance can have ethical implications. When you accept an offer, you are committing to work with that prof, who then might stop looking for other students. So by retracting your offer, you are breaking your commitment and also causing a great deal of inconvenience to the prof and also taking away the opportunity from other students. 
Ultimately, this choice is personal, involving a balance between personal benefit and the ethical impact on others.

@@

\section{Costly Gifts}\label{sec:gifts} In many countries, it is customary to give professors costly gifts (e.g., expensive liquors, jewelry, or even money), often during the holidays, to show respect and appreciation.  In the US, this is can be considered \emph{inappropriate} and discouraged. Most international students are aware of this, however, they might take it to the extreme and
